## Example code snippets

### Get and join data

```{r}
library(magrittr)
library(covid19germany)

# Get RKI data and transform to daily time series, e.g. per "Bundesland" and "Landkreis"
rki <- get_RKI_timeseries()
rki_timeseries_bundesland <- rki %>% group_RKI_timeseries("Bundesland")
rki_timeseries_landkreis <- rki %>% group_RKI_timeseries("Landkreis")

# Join population info to RKI data
rki_timeseries_bundesland <- rki_timeseries_bundesland %>%
  dplyr::left_join(ew_laender, by = "Bundesland")

rki_timeseries_landkreis <- rki_timeseries_landkreis %>%
  dplyr::left_join(ew_kreise, by = "IdLandkreis")

# Join hospital info to RKI data
rki_timeseries_bundesland <- rki_timeseries_bundesland %>%
  dplyr::left_join(hospital_beds, by = "Bundesland")
```

### Simple plots

```{r}
library(magrittr)
library(covid19germany)

get_RKI_timeseries() %>% plot_RKI_timeseries("Age", "CumNumberDead")
```

```{r}
library(ggplot2)
library(magrittr)
library(covid19germany)

rki <- get_RKI_timeseries(cache=F)

group_RKI_timeseries(rki, Bundesland) %>%
  dplyr::filter(Date > "2020-02-25") %>%
  tidyr::drop_na(Bundesland) %>%
  ggplot() +
  geom_bar(mapping = aes(x = Date,
                         y = NumberNewTestedIll,
                         fill = Bundesland),
           stat = 'identity') +
  theme_minimal() +
  ggtitle("Gemeldete Infektionen (täglich)") +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

```{r}
library(ggplot2)
library(magrittr)
library(covid19germany)

rki <- get_RKI_timeseries(cache=F)

group_RKI_timeseries(rki, Bundesland) %>%
  dplyr::filter(Date > "2020-02-25") %>%
  tidyr::drop_na(Bundesland) %>%
  dplyr::group_by(Bundesland) %>%
  dplyr::mutate(kum_fall = cumsum(NumberNewTestedIll)) %>%
  dplyr::ungroup() %>%
  ggplot() +
  geom_area(mapping = aes(x = Date,
                          y = kum_fall,
                          fill = Bundesland),
            stat = 'identity',
            na.rm = T) +
  theme_minimal() +
  ggtitle("Gemeldete Infektionen (kumulativ)") +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

Since we have the inhabitant numbers right in the package, co-analysing them toegther with the epidemiologocal data is straight forward:
  
```{r}
library(ggplot2)
library(magrittr)
library(covid19germany)

rki <- get_RKI_timeseries(cache=F)

group_RKI_timeseries(rki, Bundesland) %>%
  dplyr::left_join(ew_laender, by="Bundesland") %>%
  dplyr::filter(Date > "2020-02-25") %>%
  tidyr::drop_na(Bundesland) %>%
  dplyr::group_by(Bundesland) %>%
  dplyr::mutate(kum_fall_per100k_ew = cumsum(NumberNewTestedIll) / PopulationTotal) %>%
  dplyr::ungroup() %>%
  ggplot() +
  geom_line(mapping = aes(x = Date,
                          y = kum_fall_per100k_ew,
                          col = Bundesland)) +
  theme_minimal() +
  ggtitle("Gemeldete Infektionen pro 100K Einwohner (kumulativ)") +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

### Simple Map

```{r}
library(sf)
library(ggplot2)
library(lubridate)
library(magrittr)
library(covid19germany)

# download rki data
rki <- get_RKI_timeseries()

# download a shapefile with geoinformation for the german Landkreise
landkreis_sf <- get_RKI_spatial("Landkreis")

# download and filter rki data to 2020-03-21
rki_202003021_landkreise <- group_RKI_timeseries(rki, "Landkreis") %>% 
  dplyr::filter(Date == as_datetime("2020-03-21"))

# merge spatial data and rki data
landkreis_sf_COVID19 <- landkreis_sf %>%
  dplyr::left_join(
    rki_202003021_landkreise, by = c("IdLandkreis")
  )

# plot
landkreis_sf_COVID19 %>%
  ggplot() +
  geom_sf(
    aes(fill = CumNumberTestedIll)
  ) +
  scale_fill_viridis_c(direction = -1) +
  theme_minimal() +
  ggtitle("Summe der gemeldeten Infektionen pro Landkreis am 21.03.2020") +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

### Simple Model

A simple modelling approach to predict the number of future cases regresses historical counts on time. These predictions assume no further interventions like social distancing! Since the spread-dynamics are exponential, we choose a simple log-linear regression approach on new cases per day (i.e. not cumulative in this analysis): 
  
```{r}
library(covid19germany)

## globals:
MIN.CASES=5
N.DAYS.FUTURE=7

## read data:
df = get_RKI_timeseries()

## aggregate case-counts by regions:
ag=aggregate(df$NumberNewTestedIll,by=list(Bundesland=df$Bundesland,Date=df$Date),sum)

## split data by regions:
ag.split=split(ag,ag$Bundesland)

## plot
par(mfrow=c(4,4))
par(cex=1.2)
for (this.ag in ag.split){
  keep=(this.ag$x>=MIN.CASES)
  this.ag=this.ag[keep,]
  this.region=unique(this.ag$Bundesland)
  
  ## create model
  this.ag = this.ag[order(this.ag$Date),]
  this.ag[,"x.log"]=log(this.ag[,"x"])
  mdl=lm("x.log ~ Date",data=this.ag)
  
  ## append N.DAYS.FUTURE to data
  n=(nrow(this.ag)+1)
  last.date=as.Date(max(this.ag$Date, na.rm = T))
  this.ag[n:(n+N.DAYS.FUTURE-1),]=NA
  this.ag[n:(n+N.DAYS.FUTURE-1),"Date"]=seq(last.date+1,by=1,length.out=N.DAYS.FUTURE)
  
  ## predict on all dates (including future dates)
  this.ag[,"x.pred"]=exp(predict(mdl,newdata=this.ag))
  
  ## plot data and prediction
  ymax=max(this.ag[,c("x.pred","x")],na.rm=TRUE)
  plot(this.ag[,"Date"],this.ag[,"x"],type="b",col="black",pch=20,main=this.region,xlab="date",ylab="cases",ylim=c(0,ymax))
  points(this.ag[,"Date"],this.ag[,"x.pred"],type="b",col="red")
}
```
